Static contract checking for Haskell

### 2011-11-06 paper.tex
Paper: Removed the expression translation and fixed some typos.
The expression translation was the identity in the paper, and is recently
also the identity in the implementation.

\def\ruleform#1{{\setlength{\fboxrule}{1.2pt}\fbox{\normalsize $#1$}}}
\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ktrans}[1]{\mathcal{K} \llbracket #1 \rrbracket}
\newcommand{\ctrans}[1]{\mathcal{C} \llbracket #1 \rrbracket}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

@@

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
\renewcommand{\min}[1]{\mbox{min}(#1)}

@@

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.\note[nc]{XXX, TODO: use $True$ and $False$ in the
  text.  The lowercase $true$ and $false$ is an artifact of the implementation, i.e. Equinox accepts TPTP format, which doesn't allow bare uppercase function symbols.  Moreover, the lower case here is inconsistent with the rest of the presentation, e.g. $Cons$ and $Nil$ above.}
\begin{equation*}
\forall xs.~ xs=Nil \implies notnull(xs) = false
\end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
\[
notnull(Nil) = false
\]? The answer is that this is an expansion of the $\etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.  I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.  In this special case we could remove the implication, but not in general.  The $a = \etrans{e_i}$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}

@@

Not really true: the definition of satisfaction for arrow contracts
$x:c_1 \to c_2$ already includes partial applications.  We could add a
special cases for each function arity, e.g.
\begin{eqnarray*}
&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c} \\
&&      :=\forall x_1, \ldots, x_n. \min{\etrans{f(x_1, \ldots, x_n)}} \\
&&\quad   \to \left( \bigvee_{i=1}^n \ctrans{x_i \notin c_i} \right)
              \lor \ctrans{f(x_1, \ldots, x_n) \in c}
\end{eqnarray*}
for functions of arity $n$, to make the claim true, but that would add
more clutter.  This may be related to my confusion where removing the
$f_{ptr}$ stuff caused problems where no higher-orderness was
involved?

@@

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \etrans{\mbox{expression}} &\to& t~(\mbox{Term}) \\
    \dtrans{def} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ktrans{\mbox{data } T = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ctrans{f \in c} &\to& \Phi~(\mbox{Set of formulae}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{trans}
\end{figure}

\subsection{$\etrans{}$ -- Expressions}
Our most basic translation is from expressions in $\hprime$ to terms
in first-order logic. Given this translation we will be able to
translate definitions, data types and contracts to first-order
formulae. It is described in figure~\ref{etrans}.

\begin{figure}
$$
\begin{array}{rcl}
\multicolumn{3}{c}{\ruleform{\etrans{e} = t}} \\[2.5mm]
 \etrans{x} &=& x\\
  \etrans{f(e_1,\dots,e_n)} &=& f(\etrans{e_1},\dots,\etrans{e_n})\\
  \etrans{K(e_1,\dots,e_n)} &=& K(\etrans{e_1},\dots,\etrans{e_n})\\
  \etrans{\bad} &=& \bad
\end{array}
$$
\caption{$\etrans{}$ -- Expression translation}
\label{etrans}
\end{figure}

@@

\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
  f(\vec{x})) \to a = \etrans{e} \\
   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& a = \etrans{e_1} \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& a = \etrans{e_n} \\%removed } here ...
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

@@

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \text{\note[nc]{??? $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{x})) \\
      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

\begin{figure*}
  $$ \ruleform{\ctrans{contract} = \Phi} $$
\begin{eqnarray*}
  \ctrans{e \in \{x \mid b \}} &=&  \min{\etrans{b[e/x]}} \land (\etrans{b[e/x]} = true \lor \etrans{e} = \unr)\\
  \ctrans{e \in x:c_1 \to c_2} &=& \forall x.~\min{\etrans{e~x}} \to (\ctrans{x \not \in c_1} \lor  \ctrans{e~x \in c_2})\\
  \ctrans{e \in \cfc} &=& \min{\etrans{e}} \land \cf{\etrans{e}}
\end{eqnarray*}

\begin{eqnarray*}
  \ctrans{e \not \in \{x \mid b \}} &=&  \min{\etrans{b[e/x]}} \land (\etrans{b[e/x]} = false \lor \etrans{e} = \bad)\\
  \ctrans{e \not \in x:c_1 \to c_2} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{e~x \not \in c_2}\\
  \ctrans{e \not \in \cfc} &=& \min{\etrans{e}} \land \lnot \cf{\etrans{e}}
\end{eqnarray*}

@@

\begin{figure*}
$$\etrans{e_1~e_2} = app(e_1,e_2)$$
$$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$
$$ \cf{f\_ptr} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{f(x_1,\dots,x_n)}$$
$$\forall f\_ptr,x.~\cf{f\_ptr} \land \cf{x} \to \cf{app(f\_ptr,x)}$$
\caption{Encoding of higher-orderness}
\label{ho-fig}
\end{figure*}

>>>>

\def\ruleform#1{{\setlength{\fboxrule}{1.2pt}\fbox{\normalsize $#1$}}}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ktrans}[1]{\mathcal{K} \llbracket #1 \rrbracket}
\newcommand{\ctrans}[1]{\mathcal{C} \llbracket #1 \rrbracket}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

@@

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\ensuremath{\mathcal{H}'}}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
\renewcommand{\min}[1]{\mbox{min}(#1)}

@@

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.\note[nc]{XXX, TODO: use $True$ and $False$ in the
  text.  The lowercase $true$ and $false$ is an artifact of the implementation, i.e. Equinox accepts TPTP format, which doesn't allow bare uppercase function symbols.  Moreover, the lower case here is inconsistent with the rest of the presentation, e.g. $Cons$ and $Nil$ above.}
\begin{equation*}
\forall xs.~ xs=Nil \implies notnull(xs) = false
\end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
\[
notnull(Nil) = false
\]? The answer is that this is an expansion of the $e = K_i(\vec{x_i}) \to a = e_i$ part of the $\dtrans{}$ rule for functions defined by cases.  I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.  In this special case we could remove the implication, but not in general.  The $a = e_i$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}

@@

Not really true: the definition of satisfaction for arrow contracts
$x:c_1 \to c_2$ already includes partial applications.  We could add a
special cases for each function arity, e.g.
\begin{eqnarray*}
&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c} \\
&&      :=\forall x_1, \ldots, x_n. \min{f(x_1, \ldots, x_n)} \\
&&\quad   \to \left( \bigvee_{i=1}^n \ctrans{x_i \notin c_i} \right)
              \lor \ctrans{f(x_1, \ldots, x_n) \in c}
\end{eqnarray*}
for functions of arity $n$, to make the claim true, but that would add
more clutter.  This may be related to my confusion where removing the
$f_{ptr}$ stuff caused problems where no higher-orderness was
involved?

@@

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \dtrans{def} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ktrans{\mbox{data } T = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ctrans{f \in c} &\to& \Phi~(\mbox{Set of formulae}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{trans}
\end{figure}

\subsection{Expressions}
Expressions are the same in \hprime and FOL, so no translation is
necessary.

@@

\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
  f(\vec{x})) \to a = e \\
   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~e = K_1(\vec{x_1}) &\to& a = e_1 \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~e = K_n(\vec{x_n}) &\to& a = e_n \\
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

@@

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{??? $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{y})) \\
      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

\begin{figure*}
  $$ \ruleform{\ctrans{contract} = \Phi} $$
\begin{eqnarray*}
  \ctrans{e \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = true \lor e = \unr)\\
  \ctrans{e \in x:c_1 \to c_2} &=& \forall x.~\min{e~x} \to (\ctrans{x \not \in c_1} \lor  \ctrans{e~x \in c_2})\\
  \ctrans{e \in \cfc} &=& \min{e} \land \cf{e}
\end{eqnarray*}

\begin{eqnarray*}
  \ctrans{e \not \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = false \lor e = \bad)\\
  \ctrans{e \not \in x:c_1 \to c_2} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{e~x \not \in c_2}\\
  \ctrans{e \not \in \cfc} &=& \min{e} \land \lnot \cf{e}
\end{eqnarray*}

@@

\begin{figure*}
$$e_1~e_2 = app(e_1,e_2)$$
$$ \forall x_1,\dots,x_n.~f(x_1,\dots,x_n) = app(app(\dots app(f\_ptr,x_1),x_2),\dots,x_n)$$
$$ \cf{f\_ptr} \tot \forall x_1,\dots,x_n.~\cf{x_1} \land \dots \land \cf{x_n} \to \cf{f(x_1,\dots,x_n)}$$
$$\forall f\_ptr,x.~\cf{f\_ptr} \land \cf{x} \to \cf{app(f\_ptr,x)}$$
\caption{Encoding of higher-orderness}
\label{ho-fig}
\end{figure*}
